---
title: "las10014_midterm"
author: "Lauren Shelby"
date: "2024-03-28"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Midterm Part 1

Creating database according to the specified schema. 

```{r}
# Loading the 'Xspecies_Arabidopsis_Rice.txt' file into the environment that contains
  # the orthology information 
rice <- read.table("Xspecies_Arabidopsis_Rice.txt")

# Assigning the correct column names to the data frame 
colnames(rice) <- c("ArabidopsisGeneID", "Relationship" ,"RiceOrthologs")
head(rice)
```

```{r}
# Loading the 'AthBiomart.txt' file that contains the gene annotations into the environment 
biomart <- read.delim("AthBiomart.txt")
head(biomart)
```

```{r}
# Loading the 'AthBIOGRID.txt' file that holds gene network interactions into the environment 
biogrid <- read.delim("AthBIOGRID.txt")

# Editing down to new data frame that just contains our desired cols 6, 7, 12, and 13 
biogrid2 <- biogrid[,c("Systematic.Name.Interactor.A", "Systematic.Name.Interactor.B", "Experimental.System", "Experimental.System.Type")]
head(biogrid2)
```


Creating the 4 different tables to go into our database first in R

```{r}
# Creating the DataType table with 2 columns: DTID (primary key), and DataType
DataType <- data.frame(matrix(c(1, 2, "Gene", "GO Term"), nrow = 2, ncol = 2))
colnames(DataType) <- c("DTID", "DataType")
DataType
```

```{r}
# Creating the Data table with 3 columns: DID (primary key), DataName, and DTID_DataType (foreign key)

# Creating the 'DataName' column that will store the 'Gene.stable.ID', 'GO.term.accession', and 'RiceOrthologs' information from the biomart and biogrid data frames 
Data <- data.frame(c(unique(biomart$Gene.stable.ID), biomart$GO.term.accession, rice$RiceOrthologs)) #taking out repeated gene ID's 
colnames(Data) <- "DataName"

# Adding in the DID column that will give each DataName a value from 1 to the total number of DataNames
Data$DID <- seq_len(nrow(Data))

# Adding in DTID_DataType using a for loop and accessing the DataType table made above to use the values of DTID as the foreign key for this column 
for (i in 1:nrow(Data)) {
  if (startsWith(Data$DataName[i], "AT")) {
    Data$DTID_DataType[i] <- 1
  }
  else if (startsWith(Data$DataName[i], "LOC")) {
    Data$DTID_DataType[i] <- 1
  }
  else if (startsWith(Data$DataName[i], "GO")){
    Data$DTID_DataType[i] <- 2
  }
}

head(Data)
tail(Data)
```



Creating DataAttributes Table by combining 3 separate data frames for the Gene, GO term, and Rice Ortholog data stored in the DataNames column of the Data table. The DataAttributes table has 4 columns: DAID (primary key), DataAttributeType, DID_Data (foreign key), and DataValue. 

```{r}
library(dplyr)
library(tidyverse)

# Extracting the Arabidopsis gene related rows from the Data table
geneData <- Data[startsWith(Data$DataName, "AT"), 1:2]

# Extracting the gene name and description values of the unique Gene ID's from biomart, then creating a column for the species value for all genes. Then adding in the corresponding DID from Data in a column DID_Data. Then performing a normalization with pivot_longer to put all of the attributes in one column, and all of the values in another 
geneAttributes <- biomart %>%
  distinct(Gene.stable.ID, .keep_all = T) %>%
  select("Gene Name" = Gene.name, "Gene Description" = Gene.description) %>%
  mutate(Species = "Arabidopsis") %>%
  mutate(DID_Data = geneData$DID) %>% #Inserting the foreign key for DID col of Data
  pivot_longer(cols = c("Gene Name", "Gene Description", "Species"), 
               names_to = "DataAttributeType", values_to = "DataValue")

head(geneAttributes)
tail(geneAttributes)

```



```{r}
# Creating a data frame for the GO term attributes from the biomart table 

# Extracting the GO term related rows from the Data table 
goData <- Data[27417:217583,] # Manually selecting GO terms by coordinates because it wouldn't select right amount any other way 

# Making a data frame with the GO term accession and it's corresponding GO term 
goAttributes <- biomart %>% 
  select("DataValue" = GO.term.name) %>%
  mutate(DataAttributeType = "Go term name") %>% 
  mutate(DID_Data = goData$DID)

head(goAttributes)
tail(goAttributes)

```

```{r}
# Creating a data frame for the Rice Ortholog attributes from the rice table 

# Extracting the GO term related rows from the Data table 
riceData <- Data[startsWith(Data$DataName, "LOC"), 1:2]

# Making a data frame with the GO term accession and it's corresponding GO term 
riceAttributes <- data.frame(riceData$DID)
colnames(riceAttributes) <- "DID_Data"
riceAttributes$DataValue <- "Rice"
riceAttributes$DataAttributeType <- "Species"

head(riceAttributes)
tail(riceAttributes)
```

```{r}
# Combining all three tables into one table called DataAttributes 

DataAttributes <- bind_rows(geneAttributes, goAttributes, riceAttributes)
DataAttributes$DAID <- seq_len(nrow(DataAttributes))
head(DataAttributes)
tail(DataAttributes)
```

Creating the Relationships table in 3 separate tables then combining 

```{r}
# Creating a table containing the GOterm2Gene relationship from the biomart table 

# Creating initial table with the Gene ID's from Data (to be replaced by their DID), the DID of each GO term, and the GOterm2Gene relationship
gotogene <- data.frame(biomart$Gene.stable.ID) %>% 
  mutate(DID_Data_1 = goData$DID) %>% 
  mutate(RelValue = "GOterm2Gene") 

# Creating an index column to make sure the DID_Data_2 stays in the right order after the mutation 
gotogene$index = seq_len(nrow(gotogene)) 

#Changing column name to what it's called in geneData for merge 
colnames(gotogene)[names(gotogene) == "biomart.Gene.stable.ID"] <- "DataName"

#Performing a merge of the arabidopsis gene ID numbers in geneData to the names present in the GO term gene interactions stored in gotogene 
gotogene <- merge(gotogene, geneData, all.x = TRUE)

#Changing column name of DID to DID_Data_Gene 
colnames(gotogene)[names(gotogene) == "DID"] <- "DID_Data_2"

# Reorder columns to ensure the "DID_Data_2" column stays in its original order
gotogene <- gotogene[order(gotogene$index), ]

# Getting rid of DataName column 
gotogene <- subset(gotogene, select = -c(DataName, index))

head(gotogene)
tail(gotogene)
```

```{r}
# Creating a table with the GeneOrthologs relationship 

# Creating an initial table with the arabidopsis gene ID's from rice table 
orthologs <- data.frame(rice$ArabidopsisGeneID)

# Adding in the corresponding DID of the rice orthologs for each arabidopsis gene 
orthologs <- cbind(orthologs, riceData$DID)
colnames(orthologs) <- c("DataName", "DID_Data_1")

# Creating relationship column 
orthologs$RelValue <- "GeneOrthologs"

# Merging the arabidopsis gene ID's with the names in the orthologs table 
orthologs <- merge(orthologs, geneData, all.x = TRUE)

# Getting rid of the DataName column 
orthologs <- subset(orthologs, select = -c(DataName))

colnames(orthologs)[names(orthologs)=="DID"] <- "DID_Data_2"

head(orthologs)
tail(orthologs)
```

```{r}
# Creating a table with the Interaction relationships 

# Two separate tables with information from biogrid to match up to the corresponding Gene ID's 
genetable1 <- data.frame(biogrid2[1:56198,1]) # wouldn't grab all of them if not specifying coords
colnames(genetable1) <- "DataName"
genetable2 <- data.frame(biogrid2$Systematic.Name.Interactor.B)
colnames(genetable2) <- "DataName"

# Adding an index column to preserve the original order
genetable1$index <- seq_len(nrow(genetable1))
genetable2$index <- seq_len(nrow(genetable2))

# Merging both tables with geneData to get the corresponding Gene ID
genetable1 <- merge(genetable1, geneData, all.x = TRUE)
genetable2 <- merge(genetable2, geneData, all.x = TRUE)

# Reordering the data frames based on the index column
genetable1 <- genetable1[order(genetable1$index), ]
genetable2 <- genetable2[order(genetable2$index), ]

# Removing the index and DataName columns 
genetable1 <- subset(genetable1, select = -c(index, DataName))
genetable2 <- subset(genetable2, select = -c(index, DataName))

# Changing the two column names 
colnames(genetable1) <- "DID_Data_1"
colnames(genetable2) <- "DID_Data_2"

# Combining the two individual tables 
interactionData <- cbind(genetable1, genetable2)
interactionData$RelValue <- "InteractionRelationship"

head(interactionData)
tail(interactionData)

```
```{r}
# Combining all the tables into one Relationships table 

Relationships <- bind_rows(gotogene, orthologs, interactionData)
Relationships$RID <- seq_len(nrow(Relationships))
head(Relationships)
tail(Relationships)
```


Writing the 3 tables to SQL 

```{r}
# Connecting to RSQlite
library(RSQLite)

drv<-dbDriver("SQLite")
con<-dbConnect(drv, "las10014_midterm.sqlite")
```

```{r}
# Checking if table exists, if it does dropping the table 
if (dbExistsTable(con, "DataType")) {
  dbSendQuery(con, "drop table DataType")
}

# Writing the DataType table 
dbSendQuery(con, "CREATE TABLE DataType (DTID INTEGER PRIMARY KEY NOT NULL, DataType VARCHAR)")

insert_DT <- dbSendQuery(con, "insert into DataType (DTID, DataType)
                          values ($DTID, $DataType)")

dbBind(insert_DT, DataType)
```

```{r}
# Checking if table exists, if it does dropping the table 
if (dbExistsTable(con, "Data")) {
  dbSendQuery(con, "drop table Data")
}

# Writing the Data table 
dbSendQuery(con, "CREATE TABLE Data (DID INTEGER PRIMARY KEY NOT NULL, DataName VARCHAR, 
DTID_DataType INTEGER, FOREIGN KEY (DTID_DataType) REFERENCES DataType(DTID))")

insert_D <- dbSendQuery(con, "insert into Data (DID, DataName, DTID_DataType)
                          values ($DID, $DataName, $DTID_DataType)")

dbBind(insert_D, Data)
```

```{r}
# Checking if table exists, if it does dropping the table 
if (dbExistsTable(con, "DataAttributes")) {
  dbSendQuery(con, "drop table DataAttributes")
}

# Writing to the DataAttributes table 
dbSendQuery(con, "CREATE TABLE DataAttributes (DAID INTEGER PRIMARY KEY NOT NULL, DID_Data INTEGER, 
DataAttributeType VARCHAR, DataValue VARCHAR, FOREIGN KEY (DID_Data) REFERENCES Data(DID))")

insert_DA <- dbSendQuery(con, "insert into DataAttributes (DAID, DID_Data, DataAttributeType, DataValue)
                          values ($DAID, $DID_Data, $DataAttributeType, $DataValue)")

dbBind(insert_DA, DataAttributes)
```

```{r}
# Checking if table exists, if it does dropping the table 
if (dbExistsTable(con, "Relationships")) {
  dbSendQuery(con, "drop table Relationships")
}

# Writing to the Relationships table 
dbSendQuery(con, "CREATE TABLE Relationships (RID INTEGER PRIMARY KEY NOT NULL, DID_Data_1 INTEGER, 
DID_Data_2 INTEGER, RelValue VARCHAR, FOREIGN KEY (DID_Data_1) REFERENCES Data(DID), FOREIGN KEY(DID_Data_2) REFERENCES Data(DID))")

insert_R <- dbSendQuery(con, "insert into Relationships (RID, DID_Data_1, DID_Data_2, RelValue)
                          values ($RID, $DID_Data_1, $DID_Data_2, $RelValue)")

dbBind(insert_R, Relationships)
```

## Part 2: getOrthologs() function 


```{r}
# Creating a function getOrthologs() where the input is a GO term and the output is a data frame of Arabidopsis genes and their Rice gene orthologs that are associated with the GO term 

getOrthologs <- function(goterm, con) {
  # Getting the DID associated with the selected goterm from the DataAttributes table 
  gotermDIDQuery <- dbGetQuery(con, paste("SELECT * FROM DataAttributes WHERE DataValue = '", goterm,"'", sep = ""))
  
  # Extracting gene DIDs associated with each GO term 
  goDIDs <- gotermDIDQuery$DID_Data
  
  # Constructing the comma-separated string of gene DIDs
  goDIDs_str <- paste(goDIDs, collapse = ", ")
  
  # Querying the Relationships table for the gene DIDs associated with each GO term 
  geneIDsQuery <- dbGetQuery(con, paste("SELECT * FROM Relationships WHERE DID_Data_1 IN (", goDIDs_str, ") AND RelValue = 'GOterm2Gene'", sep = ""))
  
  # Extracting gene DIDs associated with the gene names attached to the GO term 
  geneDIDs <- geneIDsQuery$DID_Data_2
  
  # Constructing comma-separated string of gene DIDs 
  geneDIDs_str <- paste(geneDIDs, collapse = ", ")
  
   # Now using the gene DIDs to get the corresponding rice orthologs from Relationship
  orthologsQuery <- dbGetQuery(con, paste("SELECT * FROM Relationships WHERE DID_Data_2 IN (", geneDIDs_str, ") AND RelValue = 'GeneOrthologs' ", sep = ""))
  
  # Separating out the DIDs for the corresponding genes
  orthologGenes <- orthologsQuery$DID_Data_2
  orthologGenes_str <- paste(orthologGenes, collapse = ", ")
  
  # Querying the Data table for the gene names 
  orthologGeneQuery <- dbGetQuery(con, paste("SELECT DataName, DID FROM Data WHERE DID IN (", orthologGenes_str, ")", sep = ""))
  orthologGenes <- data.frame(orthologGenes)
  colnames(orthologGenes) <- "DID"
  orthologGenes$index <- seq_len(nrow(orthologGenes))
  
  # Merging the DataNames with the original table of gene DIDs that correspond to the orthologs
  orthologGeneNames <- merge(orthologGenes, orthologGeneQuery)
  orthologGeneNames <- orthologGeneNames[order(orthologGeneNames$index), ]
  
  # Separating out the DIDs for the rice orthologs from the orthologsQuery
  orthologRice <- orthologsQuery$DID_Data_1
  orthologRice_str <- paste(orthologRice, collapse = ", ")
  
  # Querying the Data table for the ortholog names 
  orthologRiceQuery <- dbGetQuery(con, paste("SELECT DataName, DID FROM Data WHERE DID IN (", orthologRice_str, ")", sep = ""))
  orthologRice <- data.frame(orthologRice)
  colnames(orthologRice) <- "DID"
  orthologRice$index <- seq_len(nrow(orthologRice))
  
  #Merging the Datanames with the original table of rice ortholog DIDs
  orthologRiceNames <- merge(orthologRice, orthologRiceQuery)
  orthologRiceNames <- orthologRiceNames[order(orthologRiceNames$index), ]
  
  # Combining the separate columns of Gene names to corresponding Rice ortholog
  Orthologs <- cbind(data.frame(orthologGeneNames$DataName, orthologRiceNames$DataName))
  colnames(Orthologs) <- c("Gene Name", "Rice Ortholog")
  
  return(Orthologs)
}

# Test the function
head(getOrthologs("binding", con))
head(getOrthologs("membrane", con))

```


```{r}
dbDisconnect(con)
```


