---
title: "las10014_midterm"
author: "Lauren Shelby"
date: "2024-03-25"
output: html_document
---

## Midterm Part 1 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Creating database according to the specified schema. 

```{r}
# Loading the 'Xspecies_Arabidopsis_Rice.txt' file into the environment that contains
  # the orthology information 
rice <- read.table("Xspecies_Arabidopsis_Rice.txt")

# Assigning the correct column names to the data frame 
colnames(rice) <- c("ArabidopsisGeneID", "Relationship" ,"RiceOrthologs")
head(rice)
```


```{r}
# Loading the 'AthBiomart.txt' file that contains the gene annotations into the environment 
biomart <- read.delim("AthBiomart.txt")
head(biomart)
```


```{r}
# Loading the 'AthBIOGRID.txt' file that holds gene network interactions into the environment 
biogrid <- read.delim("AthBIOGRID.txt")

# Editing down to new data frame that just contains our desired cols 6, 7, 12, and 13 
biogrid2 <- biogrid[,c("Systematic.Name.Interactor.A", "Systematic.Name.Interactor.B", "Experimental.System", "Experimental.System.Type")]
head(biogrid2)
```


```{r}
# Connecting to RSQlite
library(RSQLite)

drv<-dbDriver("SQLite")
con<-dbConnect(drv, "las10014_midterm.sqlite")
```

Creating the 4 different tables to go into our database first in R

```{r}
# Creating the DataType table with 2 columns: DTID (primary key), and DataType
DataType <- data.frame(matrix(c(1, 2, "Gene", "GO Term"), nrow = 2, ncol = 2))
colnames(DataType) <- c("DTID", "DataType")
DataType
```

```{r}
# Creating the Data table with 3 columns: DID (primary key), DataName, and DTID_DataType (foreign key)

# Creating the 'DataName' column that will store the 'Gene.stable.ID', 'GO.term.accession', and 'RiceOrthologs' information from the biomart and biogrid data frames 
Data <- data.frame(c(unique(biomart$Gene.stable.ID), biomart$GO.term.accession, rice$RiceOrthologs)) #taking out repeated gene ID's 
colnames(Data) <- "DataName"

# Adding in the DID column that will give each DataName a value from 1 to the total number of DataNames
Data$DID <- seq_len(nrow(Data))

# Adding in DTID_DataType using a for loop and accessing the DataType table made above to use the values of DTID as the foreign key for this column 
for (i in 1:nrow(Data)) {
  if (startsWith(Data$DataName[i], "AT")) {
    Data$DTID_DataType[i] <- 1
  }
  else if (startsWith(Data$DataName[i], "LOC")) {
    Data$DTID_DataType[i] <- 1
  }
  else if (startsWith(Data$DataName[i], "GO")){
    Data$DTID_DataType[i] <- 2
  }
}

head(Data)
tail(Data)
```

Creating DataAttributes Table by combining 3 separate data frames for the Gene, GO term, and Rice Ortholog data stored in the DataNames column of the Data table. The DataAttributes table has 4 columns: DAID (primary key), DataAttributeType, DID_Data (foreign key), and DataValue. 

```{r}
# Creating a data frame for the information with gene attributes from the biomart table 

geneAttributes <- data.frame(unique(biomart$Gene.stable.ID), unique(biomart$Gene.name), unique(biomart$Gene.description))
colnames(geneAttributes) <- c("DID_Data", "Gene Name", "Gene Description")

# Adding column for species 
geneAttributes$Species <- "Arabidopsis" 

# Performing a normalization to create the columns we want in our final data frame 
geneAttributes <- pivot_longer(geneAttributes, cols = c("Gene Name", "Gene Description", "Species"), names_to = "DataAttributeType", values_to = "DataValue")
head(geneAttributes)

# Extracting the Arabidopsis gene related rows from the Data table 
geneData <- Data[startsWith(Data$DataName, "AT"),]
head(goData)


```

```{r}
# Creating a data frame for the GO term attributes from the biomart table 

# Making a data frame with the GO term accession and it's corresponding GO term 
goAttributes <- data.frame(biomart$GO.term.accession, biomart$GO.term.name)
colnames(goAttributes) <- c("DID_Data", "DataValue") # The GO term accession will be replaced by the PK value from Data DID column 

# Adding in a column for DataAttributeType 
goAttributes$DataAttributeType <- "GO term name"

# Extracting the GO term related rows from the Data table 
goData <- Data[startsWith(Data$DataName, "GO"),]
head(goData)



goAttributes <- cbind(goAttributes, goData$DID)
head(goAttributes)


```


```{r}
# Creating a data frame for the Rice Ortholog attributes from the rice table 
riceAttributes <- data.frame(rice$RiceOrthologs)
colnames(goAttributes) <- "DID_Data" # To be replaced with PK from DID col in Data

# Adding in a column for DataAttributeType 
riceAttributes$DataAttributeType <- "Species"

# Adding in a column for DataValue
riceAttributes$DataValue <- "Rice"

# Altering DID_Data so that it is a foreign key for the DID column of Data

```


```{r}
for (i in 1:nrow(Data)) {
  if (startsWith(Data$DataName[i], "GO")) {
    for (j in 1:nrow(goAttributes)) {
      if (Data$DataName[i] == goAttributes$DID[j]) {
        goAttributes$DID[j] <- Data$DID[i]
      }
    }
  }
}
head(goAttributes)


# Creating the DataAttributes table with 4 columns: DAID (primary key), DataAttributeType, DID_Data (foreign key), and DataValue

# Initializing an empty data frame with preliminary columns to fill in 
DataAttributes <- data.frame(DataAttributeType = character(), DID_Data = integer(), DataValue = character(), stringsAsFactors = F)

# Creating a for loop to parse through each row of Data and extract the attribute type and value to fill in the DataAttribute Type and DataValue columns 

# Enter in "Arabidopsis" "Rice" and "GO term name" manually
# For Gene, perform normalizations from the biomart table 

counter <- 1 # Initializing a counter variable to keep the index of the DataAttributes table we're filling in 

for (i in 1:nrow(Data)) {
  if (startsWith(Data$DataName[i], "AT")) {
    
    
    GeneAttributes <- data.frame(matrix(c(Data$DataName[i])))
    counter <- counter + 3
  }
  else if (startsWith(Data$DataName[i], "LOC")) {
    
    counter <- counter + 1 
  }
  else if (startsWith(Data$DataName[i], "GO")) {
    
    counter <- counter + 1 
  }
}
```


```{r}
# Creating the DataAttributes table with 4 columns: DAID (primary key), DataAttributeType, DID_Data (foreign key), and DataValue 

# Creating variables that hold all of the attribute types 
Attributes_Gene <- c("Gene.name", "Gene.description", "Species")
Attributes_GO <- "GO.term.name"

# Initializing an empty data frame with correct column names 
DataAttributes <- data.frame(DAID = integer(), DataAttributeType = character(), DID_Data = integer(), DataValue = character(), stringsAsFactors = F)

# Creating a for loop to fill in the DataAttributeType and DID_Data for each gene and GO term

for (i in 1:nrow(Data)) { # Parsing through each row of Data to access each gene or GO term 
  if (Data$DTID_DataType[i] == 1){ # If the row in Data is a Gene 
    for (j in 0:length(Attributes_Gene)) {
      DataAttributes$DataAttributeType[i+j] <- Attributes_Gene[j+1]
      DataAttributes$DID_Data[i+j] <- Data$DID[i]
    }
  }
  else if (Data$DTID_DataType[i] == 2) { # If the row in Data is a GO term
    DataAttributes$DataAttributeType[i,2] <- Attributes_GO
  }
}

# Creating a for loop to assign the value DataValue column in each column of DataAttributes

# Creating a for loop to fill in the DataAttribute Type information for each GO term 


colnames(DataAttributes) <- "DataAttributeType"

# Creating a column DAID that holds the primary key for this table going 1 through nrow(DataAttributes)
DataAttributes$DAID <- seq_len(nrow(DataAttributes))

```


```{r}
# Creating the Relationship table with 5 columns: RID (primary key), Relationship, DID_Data_1 (foreign key), DID_Data_2 (foreign key), and RelValue

```

```{r}
dbDisconnect(con)
```

